Storm surges in the Strait of Georgia simulated with a regional model

### 2015-06-08 AO_guides/manuscriptAO.tex
Susan's comments and some of Keith's wordy comments

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its skill in simulating storm surge hindcasts. {\color{red} The main purposes are to identify the forcing factors that contribute most significantly to storm surges in the Strait of Georgia, to examine the spatial distribution of the storm surge, and to evaluate the performance of a new model of this region.}  To complement previous barotropic storm surge modelling in this region, this model employs a higher resolution and includes the effect of stratification and rivers. However, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain is provided in section \ref{sec:config}. That section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's skill in producing storm surge hindcasts is assessed in section \ref{sec:storm}. Several simulations comparing the effects of different forcing conditions are also discussed. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait (Figure \ref{fig:domain}). Bathymetry from the Cascadia physiography data set \citep{haugerud1999digital} was smoothed to limit large changes in depth across grid cells, so that $\Delta h/\bar{h} \leq 0.8$, where $\Delta h$ is the difference in depth between two adjacent grid cells and $\bar{h}$ is their average depth. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells {\color{red} inward from the open boundary. Depths between 0 m and 4 m were set to 4 m and a minimum depth of 4 m was imposed by NEMO.}  As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis. {\color{red} A magnified image of the complex passages through the San Juan and Gulf Islands is provided in Figure \ref{fig:zoom}.}

@@

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results of a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid to give a two-dimensional forcing that varies with time.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5 km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. {\color{red} At the north, seasonal climatologies from \citet{thomson1981oceanography} were used in the same manner.} %other references?

Sea surface height anomaly (or remote forcing) was set uniformly across the mouth of Juan de Fuca using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain (Figure \ref{fig:domain}).  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. {\color{red}Tidal predictions were generated with the tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year of the storm surge. A harmonic analysis was performed for 67 constituents. In order to prevent over-predicted constituents due to high surge years, the time series was first filtered by applying a Doodson filter \citep{parker2007NOAA} and removing periods with large non-tidal amplitudes. Long period constituents such as SA, SSA, etc were not used in the tidal predictions since these constituents are often contaminated by non-tidal energy from seasonal meteorological trends. Further, constituents with a low signal to noise ratio were neglected. } The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait.

{\color{red} Although Tofino is not the closest tide guage to the western boundary, we have found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. In the future, storm surge forecasts will make use of Neah Bay water level predictions from the National Oceanic and Atmospheric Administration (NOAA).}

\begin{table}[h]
\centering 
\tbl{{\color{red}Average surge amplitudes and delay in peak timing from Tofino at several tide gauges on the Pacific coast. Averages obtained over the surge events discussed on this article. A negative delay indicates that the peak surge occurred later than at Tofino.}} 
{\begin{tabular}{|l |c | c|}
\hline 
Location & Mean surge amplitude (m) & Mean delay from Tofino (hrs) \\
\hline
Tofino  & 0.707 & -- \\
Toke Point & 1.17 & -2.83 \\
Neah Bay & 0.658 & -1.83 \\
Bamfield & 0.658 & -0.5 \\
Winter Harbour & 0.572 & 0.333 \\
Port Hardy & 0.491 & -2.67\\
\hline
\end{tabular}}
\label{tab:pacificsurge}
\end{table}

>>>>

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its skill in simulating storm surge hindcasts. {\color{red} The main goals are to identify the forcing factors that contribute most significantly to storm surges in the Strait of Georgia, to examine the spatial distribution of the storm surges, and to evaluate the performance of a new model of this region.}  To complement previous barotropic storm surge modelling in this region, this model employs a higher resolution and includes the effect of stratification and rivers. However, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain is provided in section \ref{sec:config}. That section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's skill in producing storm surge hindcasts is assessed in section \ref{sec:storm}. Several simulations comparing the effects of different forcing conditions are also discussed. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait (Figure \ref{fig:domain}). Bathymetry from the Cascadia physiography data set \citep{haugerud1999digital} was smoothed to limit large changes in depth across grid cells, so that $\Delta h/\bar{h} \leq 0.8$, where $\Delta h$ is the difference in depth between two adjacent grid cells and $\bar{h}$ is their average depth. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells {\color{red} inward from the open boundary. Depths between 0 m and 4 m were set to 4 m.}  As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis. {\color{red} A magnified image of the complex passages through the San Juan and Gulf Islands is provided in Figure \ref{fig:zoom}.}

@@

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results of a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid to give a two-dimensional forcing that varies with time.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5 km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. {\color{red} In the north, the seasonal climatologies of \citet{thomson1981oceanography} were used in the same manner.} %other references?

Sea surface height anomaly (or remote forcing) was set uniformly across the mouth of Juan de Fuca using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain (Figure \ref{fig:domain}).  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. {\color{red}Tidal predictions were generated with the tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year of the storm surge. A harmonic analysis was performed for 67 constituents. In order to prevent over-predicted constituents due to high surge years, the time series was first filtered by applying a Doodson filter \citep{parker2007NOAA} and periods with large non-tidal amplitudes were removed. Long period constituents such as Sa, Ssa, etc were not used in the tidal predictions since these constituents are often contaminated by non-tidal energy from seasonal meteorological trends. Further, constituents with a low signal to noise ratio were neglected. } The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait.

{\color{red} Although Tofino is not the closest tide gauge to the western boundary, we found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. In the future, storm surge forecasts will make use of Neah Bay water level predictions from the National Oceanic and Atmospheric Administration (NOAA).}

\begin{table}[h]
\centering 
\tbl{{\color{red}Average surge amplitudes and delay in peak timing from Tofino at several tide gauges on the Pacific coast. Averages obtained over the surge events discussed on this article. A negative delay indicates that the peak surge occurred later than at Tofino.}} 
{\begin{tabular}{|l |c | c|}
\hline 
Location & Mean surge amplitude (m) & Mean delay from Tofino (hrs) \\
\hline
Port Hardy & 0.491 & -2.67\\
Winter Harbour & 0.572 & 0.333 \\
Tofino  & 0.707 & -- \\
Bamfield & 0.658 & -0.5 \\
Neah Bay & 0.658 & -1.83 \\
Toke Point & 1.17 & -2.83 \\
\hline
\end{tabular}}
\label{tab:pacificsurge}
\end{table}

{\color{red} The remote forcing is applied uniformly across the open boundaries. Given the similarities between the surge amplitude and phase at Neah Bay and Bamfield, this seems to be a reasonable assumption. Further, a geostrophic slope calculation with velocity 20 cm s$^{-1}$ would imply a difference of less than 5 cm between the northern and southern sides of the Strait of Juan de Fuca. Non-uniform forcing conditions will be investigated in a future study.}
